---
title: "Prenatal Glyphosate, AMPA, and Oxidative Stress Biomarkers"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme: cosmo
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(shiny)
library(flexdashboard)
library(dplyr)
library(tidyr)
library(broom)
library(forcats)
library(plotly)
library(htmltools)
library(readxl)
library(tidyverse)
library(stringr)
library(EnvStats)
```

```{css, echo=FALSE}
.main-container, .container-fluid {
  max-width: 100% !important;
  width: 100% !important;
}

body {
  font-size: 16px;
}

.navbar-brand {
  font-weight: 700;
}

.section.level3 {
  overflow-x: auto;
}

.plotly.html-widget {
  max-width: 100%;
}

.small-note {
  font-size: 0.95rem;
  line-height: 1.45;
  margin-top: 0.75rem;
}

.impact-box {
  padding: 1rem;
  border-radius: 10px;
  background: #f7f7f7;
  border-left: 4px solid #555;
  margin-top: 1rem;
}

.filter-container {
  display: flex;
  justify-content: center;
  gap: 1rem;
  margin: 0.15rem 0 0.25rem 0;
  flex-wrap: wrap;
}

.filter-box-small {
  width: 230px;
}

.filter-box-small label {
  font-size: 12px;
  margin-bottom: 2px;
}

.filter-box-small .form-control {
  height: 34px;
  font-size: 13px;
  padding: 3px 8px;
}

.chart-wrapper {
  margin-bottom: 6px;
}

.takeaway-box {
  margin-top: 4px;
  padding: 12px 16px;
  background-color: #f8f9fa;
  border-left: 5px solid #2c7fb8;
  border-radius: 6px;
  font-size: 14px;
  line-height: 1.4;
}

@media (max-width: 768px) {
  body { font-size: 14px; }
  .chart-title { font-size: 1.1rem; }
  .control-label { font-size: 0.9rem; }
}
@media (max-width: 768px) {
  .filter-box-small {
    width: 90%;
  }

  .chart-wrapper {
    overflow-x: auto;
    -webkit-overflow-scrolling: touch;
  }

  .plotly.html-widget {
    min-width: 850px;
  }

  .takeaway-box {
    font-size: 13px;
  }
}
@media (max-width: 768px) {

  /* shrink facet titles (biomarker names) */
  .plotly .gtitle {
    font-size: 12px !important;
  }

  /* shrink axis labels */
  .plotly .xtick text,
  .plotly .ytick text {
    font-size: 10px !important;
  }

  /* reduce subplot spacing */
  .plotly .subplot {
    padding: 5px !important;
  }

  /* give more bottom space so x-axis doesn't overlap */
  .plotly .cartesianlayer {
    margin-bottom: 20px !important;
  }
}
.mobile-scroll-chart {
  width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
  -webkit-overflow-scrolling: touch;
}

.mobile-scroll-chart .plotly {
  min-width: 1050px;
}
```

```{r clean-data, include=FALSE}
# Simulated dataset included in GitHub repository.
merged <- readxl::read_excel("simulated_data.xlsx")

set.seed(555)

# Simulate AMPA using the original AMPA geometric mean and geometric SD.
# Original AMPA + 0.0001 summary:
# geometric mean = 0.02968493
# geometric SD = 7.636166
merged <- merged |>
  mutate(
    ampa_constant_sim = exp(
      rnorm(
        n(),
        mean = log(0.02968493),
        sd = log(7.636166)
      )
    ),
    AMPA_PA_SG.sim = pmax(ampa_constant_sim - 0.0001, 0)
  )

merged <- merged |>
  mutate(
    mat_age = mat_age.sim,
    ppbmi = ppbmi.sim,
    mat_educ = mat_educ.sim,
    AHEI_WGRAINS_PA = ahei.sim,
    parity = parity.sim,
    marital = marital.sim,
    mat_race_eth = mat_race_eth.sim,
    employ = employ.sim,

    # Recoding
    marital = if_else(marital == 3, 2, marital),
    mat_race_eth = case_when(
      mat_race_eth %in% c(3, 4, 5) ~ 2,
      TRUE ~ mat_race_eth
    ),

    # Oxidative stress biomarkers
    log.iso_sg_avg = log(iso_sg_avg.sim),
    log.isom_sg_avg = log(isom_sg_avg.sim),
    log.isom_new_sg_avg = log(isom_new_sg_avg.sim),
    log.pgf_sg_avg = log(pgf_sg_avg.sim),
    log.iso_chem_sg_avg = log(iso_chem_sg_avg.sim),
    log.iso_enz_sg_avg = log(iso_enz_sg_avg.sim),

    # Exposures
    Glyphosate_PA_SG = Glyphosate_PA_SG.sim,
    AMPA_PA_SG = AMPA_PA_SG.sim,
    log_Glyphosate_PA_SG = log(Glyphosate_PA_SG),
    ampa_constant = AMPA_PA_SG + 0.0001,
    log_AMPA_PA_SG = log(ampa_constant),

    # IQR-scaled exposures
    glyph.iqr = log_Glyphosate_PA_SG / IQR(log_Glyphosate_PA_SG, na.rm = TRUE),
    ampa.iqr = log_AMPA_PA_SG / IQR(log_AMPA_PA_SG, na.rm = TRUE)
  )

merged$glyph_tertile <- cut(
  merged$Glyphosate_PA_SG,
  breaks = quantile(
    merged$Glyphosate_PA_SG,
    probs = c(0, 1/3, 2/3, 1),
    na.rm = TRUE
  ),
  include.lowest = TRUE,
  labels = c("Low", "Medium", "High")
)

merged$ampa_tertile <- cut(
  merged$AMPA_PA_SG,
  breaks = quantile(
    merged$AMPA_PA_SG,
    probs = c(0, 1/3, 2/3, 1),
    na.rm = TRUE
  ),
  include.lowest = TRUE,
  labels = c("Low", "Medium", "High")
)

merged$glyph_tertile <- relevel(merged$glyph_tertile, ref = "Low")
merged$ampa_tertile  <- relevel(merged$ampa_tertile, ref = "Low")
```

```{r helper-functions, include=FALSE}
rename_outcomes <- function(df) {
  df %>%
    mutate(
      outcome = recode(
        outcome,
        "log.iso_sg_avg"      = "8-iso-PGF2α",
        "log.isom_sg_avg"     = "2,3-dinor-5,6-dihydro-8-iso-PGF2α",
        "log.isom_new_sg_avg" = "2,3-dinor-8-iso-PGF2α",
        "log.pgf_sg_avg"      = "PGF2α"
      )
    )
}

run_exposure_results <- function(data, exposure, exposure_label, tertile_labels) {
  outcome_vars <- c(
    "log.iso_sg_avg",
    "log.isom_sg_avg",
    "log.isom_new_sg_avg",
    "log.pgf_sg_avg"
  )

  run_iqr <- function(outcome_var) {
    fit <- lm(
      as.formula(
        paste0(
          outcome_var,
          " ~ ", exposure, ".iqr + mat_age + ppbmi + AHEI_WGRAINS_PA + as.factor(mat_educ)"
        )
      ),
      data = data
    )

    tidy(fit, conf.int = TRUE) %>%
      filter(term == paste0(exposure, ".iqr")) %>%
      mutate(
        outcome = outcome_var,
        metric = "IQR increase",
        exposure_type = exposure_label
      )
  }

  run_tert <- function(outcome_var) {
    fit <- lm(
      as.formula(
        paste0(
          outcome_var,
          " ~ ", exposure, "_tertile + mat_age + ppbmi + AHEI_WGRAINS_PA + as.factor(mat_educ)"
        )
      ),
      data = data
    )

    tidy(fit, conf.int = TRUE) %>%
      filter(grepl(paste0("^", exposure, "_tertile"), term)) %>%
      mutate(
        outcome = outcome_var,
        exposure_type = exposure_label
      )
  }

  iqr_res <- bind_rows(lapply(outcome_vars, run_iqr))
  tert_res <- bind_rows(lapply(outcome_vars, run_tert))

  tert_clean <- tert_res %>%
    mutate(
      metric = case_when(
        grepl("Medium$", term) ~ tertile_labels[2],
        grepl("High$", term)   ~ tertile_labels[3],
        TRUE                   ~ NA_character_
      )
    )

  ref_rows <- lapply(outcome_vars, function(outcome_var) {
    tibble(
      term = paste0(exposure, "_tertileLow"),
      estimate = 0,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      conf.low = 0,
      conf.high = 0,
      outcome = outcome_var,
      metric = tertile_labels[1],
      exposure_type = exposure_label
    )
  }) %>% bind_rows()

  bind_rows(iqr_res, tert_clean, ref_rows) %>%
    rename_outcomes() %>%
    mutate(
      outcome = factor(
        outcome,
        levels = c(
          "8-iso-PGF2α",
          "2,3-dinor-5,6-dihydro-8-iso-PGF2α",
          "2,3-dinor-8-iso-PGF2α",
          "PGF2α"
        )
      ),
      metric = factor(metric, levels = c("IQR increase", tertile_labels))
    )
}

glyph_q <- quantile(merged$Glyphosate_PA_SG, probs = c(1/3, 2/3), na.rm = TRUE)
ampa_q  <- quantile(merged$AMPA_PA_SG, probs = c(1/3, 2/3), na.rm = TRUE)

glyph_labels <- c(
  paste0("T1: ≤ ", round(glyph_q[1], 3), " ng/mL"),
  paste0("T2: ", round(glyph_q[1], 3), "–", round(glyph_q[2], 3), "<br>ng/mL"),
  paste0("T3: > ", round(glyph_q[2], 3), " ng/mL")
)

ampa_labels <- c(
  paste0("T1: ≤ ", round(ampa_q[1], 3), " ng/mL"),
  paste0("T2: ", round(ampa_q[1], 3), "–", round(ampa_q[2], 3), "<br>ng/mL"),
  paste0("T3: > ", round(ampa_q[2], 3), " ng/mL")
)

glyph_results <- run_exposure_results(merged, "glyph", "Glyphosate", glyph_labels)
ampa_results <- run_exposure_results(merged, "ampa", "AMPA", ampa_labels)

plotdf_all <- bind_rows(glyph_results, ampa_results) %>%
  mutate(
    hover_text = paste0(
      "<b>Exposure:</b> ", exposure_type,
      "<br><b>Biomarker:</b> ", outcome,
      "<br><b>Metric:</b> ", metric,
      "<br><b>Beta:</b> ", sprintf("%.3f", estimate),
      "<br><b>95% CI:</b> ", sprintf("%.3f", conf.low), " to ", sprintf("%.3f", conf.high)
    )
  )
```

Overview {data-navmenu="About"}
=====================================

Row
-------------------------------------

### Dashboard guide

Use the [**Adjusted Associations**](#adjusted-associations) tab to explore linear regression results adjusted for age, diet, pre-pregnancy BMI, and education for each exposure and biomarker. You can filter by exposure type to compare patterns across biomarkers and hover over points to view the exact beta coefficient and confidence interval. Use the [**Correlation Heatmap**](#correlation-heatmap) tab to compare pairwise Pearson correlations among glyphosate, AMPA, and oxidative stress biomarkers. Hover over each figure for rounded estimates and labels.

The source code can be found here: https://github.com/jarwari/GlyphosateDashboard/tree/main

### Dataset description

This dashboard uses simulated data modeled after the Illinois Kids Development Study (I-KIDS), a prospective birth cohort examining prenatal environmental exposures and child health outcomes. The simulated analytic dataset includes 161 participants with linked urinary measurements of glyphosate and AMPA collected during early pregnancy and oxidative stress biomarkers measured in the second and/or third trimester. The original confidential data were collected between 2012 and 2017 from pregnant individuals in Champaign-Urbana, Illinois, with biospecimens obtained through clinical collection and analyzed using laboratory-based methods. Biomarkers were specific gravity corrected and natural log transformed prior to analysis. The original data are confidential and not publicly available.

### Why this matters

Glyphosate is widely used, yet human evidence on prenatal glyphosate and AMPA exposure in relation to oxidative stress during pregnancy remains limited. This dashboard helps communicate whether exposure patterns are associated with oxidative stress biomarkers and supports continued investigation into potential biological effects during a sensitive developmental window.

Adjusted Associations
=====================================

Column {data-width=1000}
-------------------------------------

### Adjusted Associations between Glyphosate or AMPA and Oxidative Stress Biomarkers

```{r controls}
div(
  class = "filter-container",
  div(
    class = "filter-box-small",
    selectInput(
      "exposure_filter",
      "Exposure",
      choices = c("Glyphosate", "AMPA"),
      selected = "Glyphosate"
    )
  ),
  div(
    class = "filter-box-small",
    selectInput(
      "biomarker_filter",
      "Biomarker",
      choices = c(
        "All",
        "8-iso-PGF2α",
        "2,3-dinor-5,6-dihydro-8-iso-PGF2α",
        "2,3-dinor-8-iso-PGF2α",
        "PGF2α"
      ),
      selected = "All"
    )
  )
)
```

```{r widget1-plot}
div(
  class = "mobile-scroll-chart",
  plotlyOutput("widget1_plot", width = "1050px", height = "500px")
)

output$widget1_plot <- renderPlotly({

  df <- plotdf_all %>%
    filter(exposure_type == input$exposure_filter)

  if (input$biomarker_filter != "All") {
    df <- df %>% filter(outcome == input$biomarker_filter)
  }

  validate(need(nrow(df) > 0, "No data available."))

  metric_levels <- if (input$exposure_filter == "Glyphosate") {
    c("IQR increase", glyph_labels)
  } else {
    c("IQR increase", ampa_labels)
  }

  df <- df %>%
    mutate(
      metric = factor(metric, levels = rev(metric_levels)),
      outcome = factor(
        outcome,
        levels = c(
          "8-iso-PGF2α",
          "2,3-dinor-5,6-dihydro-8-iso-PGF2α",
          "2,3-dinor-8-iso-PGF2α",
          "PGF2α"
        )
      )
    )

  line_color <- ifelse(input$exposure_filter == "Glyphosate", "#1b9e77", "#d95f02")

  p <- ggplot(
    df,
    aes(
      x = estimate,
      y = metric,
      text = hover_text
    )
  ) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    geom_errorbarh(
      aes(xmin = conf.low, xmax = conf.high),
      height = 0.15,
      linewidth = 0.8,
      color = line_color
    ) +
    geom_point(size = 3, color = line_color) +
    facet_wrap(~ outcome, nrow = 1, scales = "free_x") +
    labs(
      title = input$exposure_filter,
      x = "Beta coefficient (95% CI)",
      y = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
      strip.text = element_text(face = "bold", size = 11),
      axis.text.y = element_text(size = 11),
      axis.text.x = element_text(size = 10),
      axis.title.x = element_text(size = 12, margin = margin(t = 12)),
      panel.spacing = unit(1.4, "lines"),
      plot.margin = margin(15, 20, 35, 20)
    )

  ggplotly(p, tooltip = "text") %>%
    layout(
      autosize = TRUE,
      margin = list(l = 90, r = 40, b = 140, t = 70)
    ) %>%
    config(responsive = FALSE, displayModeBar = TRUE)
})
```

<div class="takeaway-box">
<strong>Key takeaways:</strong> Using the simulated dataset, there were some negative, significant associations between AMPA and 2,3-dinor-5,6-dihydro-8-iso-PGF2α; however, this association is not present while using the original dataset. Overall, the forest plots suggest limited evidence of consistent associations between glyphosate or AMPA exposure and oxidative stress biomarkers, with most estimates centered near the null and wide confidence intervals. While some variability is observed across exposure tertiles and biomarkers, findings are largely null, supporting the need for continued investigation into potential biological effects during pregnancy. 
</div>

Correlation Heatmap
=====================================

Column {data-width=1000}
-------------------------------------

### Pearson correlations among exposures and oxidative stress biomarkers

```{r widget2-data, include=FALSE}
corr_data <- merged %>%
  select(
    Glyphosate = log_Glyphosate_PA_SG,
    AMPA = log_AMPA_PA_SG,
    `8-iso-PGF2α` = log.iso_sg_avg,
    `2,3-dinor-5,6-dihydro-8-iso-PGF2α` = log.isom_sg_avg,
    `2,3-dinor-8-iso-PGF2α` = log.isom_new_sg_avg,
    PGF2α = log.pgf_sg_avg
  ) %>%
  filter(complete.cases(.))

corr_matrix <- cor(corr_data, method = "pearson")
var_names <- colnames(corr_matrix)
n_vars <- length(var_names)

x_labels_wrapped <- c(
  "Glyphosate",
  "AMPA",
  "8-iso-PGF2α",
  "2,3-dinor-5,6-dihydro-<br>8-iso-PGF2α",
  "2,3-dinor-8-iso-PGF2α",
  "PGF2α"
)

get_strength <- function(r) {
  abs_r <- abs(r)
  dplyr::case_when(
    abs_r >= 0.80 ~ "Very strong",
    abs_r >= 0.60 ~ "Strong",
    abs_r >= 0.40 ~ "Moderate",
    abs_r >= 0.20 ~ "Weak",
    TRUE ~ "Very weak"
  )
}

z_mat <- corr_matrix
text_mat <- matrix("", nrow = n_vars, ncol = n_vars)

for (i in seq_len(n_vars)) {
  for (j in seq_len(n_vars)) {
    if (i < j) {
      z_mat[i, j] <- NA
      text_mat[i, j] <- ""
    } else {
      r_val <- corr_matrix[i, j]
      text_mat[i, j] <- paste0(
        "<b>Variable 1:</b> ", var_names[j],
        "<br><b>Variable 2:</b> ", var_names[i],
        "<br><b>Pearson r:</b> ", sprintf("%.2f", r_val),
        "<br><b>Relationship strength:</b> ", get_strength(r_val)
      )
    }
  }
}

z_plot <- z_mat[nrow(z_mat):1, ]
text_plot <- text_mat[nrow(text_mat):1, ]
y_labels <- rev(var_names)
x_labels <- var_names
```

```{r widget2-plot}
div(
  class = "chart-wrapper",
  plotlyOutput("widget2_plot", width = "100%", height = "500px")
)

output$widget2_plot <- renderPlotly({
  plot_ly(
    x = x_labels,
    y = y_labels,
    z = z_plot,
    text = text_plot,
    type = "heatmap",
    hovertemplate = "%{text}<extra></extra>",
    zmin = -1,
    zmax = 1,
    showscale = TRUE,
    colorbar = list(
      title = list(text = "Pearson r", side = "top"),
      tickvals = c(-1, -0.5, 0, 0.5, 1),
      ticktext = c("-1.0", "-0.5", "0", "0.5", "1.0"),
      len = 0.70,
      thickness = 18,
      x = 1.03
    ),
    xgap = 2,
    ygap = 2
  ) %>%
    layout(
      title = list(
        text = "<b>Pearson Correlations Among Exposures and Oxidative Stress Biomarkers</b>",
        x = 0.5,
        xanchor = "center",
        font = list(size = 18)
      ),
      xaxis = list(
        title = "",
        side = "bottom",
        tickmode = "array",
        tickvals = x_labels,
        ticktext = x_labels_wrapped,
        tickangle = 0,
        tickfont = list(size = 12),
        automargin = TRUE
      ),
      yaxis = list(
        title = "",
        tickfont = list(size = 13),
        automargin = TRUE
      ),
      margin = list(l = 220, r = 90, b = 115, t = 80),
      paper_bgcolor = "white",
      plot_bgcolor = "white"
    ) %>%
    config(responsive = TRUE, displayModeBar = TRUE)
})
```

<div class="takeaway-box">
<strong>Key takeaways:</strong> The heatmap suggests that correlations among glyphosate, AMPA, and oxidative stress biomarkers are weak in magnitude, with no clear evidence of very strong pairwise relationships.
</div>